---
title: "2022DMF1C Asia"
author: "Luis Fernando Delgado"
date: "02/02/2023"
---

# Introduction

This is a script about predictions in non-replicated trials with augmented checks. According to Salvador Gezan, the best way to make good predictions is to do a spatial analysis. See [Part 1](https://vsni.co.uk/blogs/unreplicated-trials-part-1) and [Part 2](https://vsni.co.uk/blogs/unreplicated-trials-part-2).

## Load packages 

Here we check whether all necessary packages are installed in your computer.

```{r}
## Packages (install)
pkgs <- rownames(installed.packages())
if(!"tidyverse" %in% pkgs) install.packages("tidyverse")
if(!"here" %in% pkgs) install.packages("here")
if(!"statgenSTA" %in% pkgs) install.packages("statgenSTA")
if(!"statgenGxE" %in% pkgs) install.packages("statgenGxE")
if(!"openxlsx" %in% pkgs) install.packages("openxlsx")
if(!"asreml" %in% pkgs) install.packages("asreml")
```

The code recognizes if certain packages are missing from your package library and, if so, will install them automatically.

Now, we need to load the packages:

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(statgenSTA)
library(statgenGxE)
library(openxlsx)
library(asreml)
```

## Load the functions of utilities_tidy on GitHub

```{r}
source("https://raw.githubusercontent.com/Cassava2050/PPD/main/utilities_tidy.R")
source("https://raw.githubusercontent.com/darizasu/work/master/scripts/ggCor.R")
```

